---
output: html_document
---
# Reproducible Research: Peer Assessment 1

First of all, load required libraries and remove warnings to avoid unwanted text in the output.
```{r}
options (warn=-1)
library(ggplot2)
library(plyr)
```

## Loading and preprocessing the data

First step, unzip the data and read the 'csv' file. Print a summary to have an idea about the data and, finally, reformat the date information from 'factor' to 'date' type.
```{r}
data <- read.csv(unzip("activity.zip"))
summary(data)
data$date <- as.Date(data$date, format = "%Y-%m-%d")
```

From now on, the questions included in the assigment are going to be included and answered.

## What is mean total number of steps taken per day?

1. *Make a histogram of the total number of steps taken each day*

To plot the histogram, the number of steps per date are stored in the variable 'steps_day'. As it is stated in the assignement, NA are ignored. Attributes are renamed to be meaningfull and, finally, the histogram plot is created using ggplot2.

```{r}
steps_day <- aggregate(data$steps, by= list(data$date), sum)
names(steps_day) <- c("date", "total_steps")
ggplot(steps_day, aes(x=total_steps)) + geom_histogram(binwidth = 1000, colour="white", fill="#FFCC66")
```

2. *Calculate and report the mean and median total number of steps taken per day*

First, a summary is included to provide a more complete idea of the data. This summay include information regarding mean and median.

```{r}
summary(steps_day)
mean_steps_day <- mean(steps_day$total_steps, na.rm = TRUE)
median_steps_day <- median(steps_day$total_steps, na.rm = TRUE)
```

Even though, to simplify the task, the mean number of steps is `r mean_steps_day` and the median is `r median_steps_day`.

## What is the average daily activity pattern?

1. *Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)*

```{r}
steps_interval <- aggregate(data$steps, by= list(data$interval), FUN=mean, na.rm=TRUE)
names(steps_interval) <- c("interval", "avg_steps")
ggplot(steps_interval, aes(x=interval, y=avg_steps)) + geom_line()
```

2. *Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?*

```{r}
max_step <- max(steps_interval$avg_steps)
steps_interval[steps_interval$avg_steps == max_step, ]
```

## Imputing missing values

1. *Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)*

No hay valores con NA en interval ni en date (¿poner prueba?)
```{r}
nrow(data[is.na(data$steps) == TRUE, ])
```

2. *Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.*

Si steps == NA entonces usar la media de los steps en ese mismo intervalo del resto de días "steps_interval"
En caso de ser 0 (mirar como asegurar que viene porque todos son null), poner el valore del intervalo posterior con datos

3. *Create a new dataset that is equal to the original dataset but with the missing data filled in.*

```{r}
data_full <- data
data_full$averages <- steps_interval$avg_steps
data_full$steps[is.na(data_full$steps)] <- 
      data_full$averages[is.na(data_full$steps)]
```

4. *Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?*

```{r}
steps_day_full <- aggregate(data_full$steps, by= list(data_full$date), sum)
names(steps_day_full) <- c("date", "total_steps")
ggplot(steps_day_full, aes(x=total_steps)) + geom_histogram(binwidth = 1000, colour="white", fill="#FFCC66")
summary(steps_day_full)
```


## Are there differences in activity patterns between weekdays and weekends?

1. *Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.*

```{r}
data_full$weekday <- ifelse(format(data_full$date, format = "%w") %in% c(0,5), "weekend", "weekday")
```


2. *Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:*

```{r}
steps_interval_wd <- aggregate(data_full$steps, by= list(data_full$interval, data_full$weekday), FUN=mean, na.rm=TRUE)
names(steps_interval_wd) <- c("interval", "weekday", "avg_steps")
ggplot(steps_interval_wd, aes(x=interval, y=avg_steps)) + facet_grid(weekday ~ .) + geom_line()
```

